---
title: "The Nobel Prize"
output: 
  flexdashboard::flex_dashboard:
    css: styles.css
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r global, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(ggplot2)
library(highcharter)
library(shiny)
library(dplyr)
library(ggplot2)
library(flexdashboard)
library(highcharter)
library(leaflet)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(sf)
library(tidyr)
library(dplyr)
library(tidyverse)
library(sf)


Nobel <- read.csv("Nobel_final_data.csv")
```

Introduction
==========================

#### Background Information

The Nobel Prize is one of the most prestigious international awards, recognizing individuals and organizations that have made major contributions to science, literature, economics, and peace. 

In this Shiny dashboard, we analyze Nobel Prize data collected from the official Nobel Prize API. The dataset includes information about laureates such as age, gender, birth country, prize category, prize amount, and year awarded.

Using this project, we explore historical trends in Nobel Prize awards across different time periods and categories. We also investigate patterns related to gender representation and geographical distribution in order to better understand inequalities and changes in recognition over time.

<<<<<<< HEAD
=======


#### Data Sources

- [Nobel Prize Website](https://www.nobelprize.org/)
- [Nobel Prize API Page](https://www.nobelprize.org/about/developer-zone-2/)
- [Nobel Prize API](https://api.nobelprize.org/2.1/laureates)

#### GitHub

Here's a link to our [code](https://github.com/AithaIriho/i-m-so-done.git)
 
Row {data-width=6}
-------------------------------------

Column {data-width=6}
-------------------------------------
>>>>>>> 11e870b921091118b04cab223700237fd46056e8
Do you know what all these following women have in common ? 

- Rosalind Franklin

- Lise Meitner

- Alice Ball 

- Jocelyn Bell Burnell

- Marthe Gautier

- Cecilia Payne-Gaposchkin

- Esther Lederberg 

- Chien-Shiung Wu

- Margaret Knight

- Hedy Lamarr

- Ada Lovelace
 
 
#### Data Sources

- [Nobel Prize Website](https://www.nobelprize.org/)
- [Nobel Prize API Page](https://www.nobelprize.org/about/developer-zone-2/)
- [Nobel Prize API](https://api.nobelprize.org/2.1/laureates)

#### GitHub

Here's a link to our [code](https://github.com/)

 
Overview
==========================

```{r, include=FALSE}

Nnobel <- Nobel |>
  group_by(Prize_year) |>
  summarise(number_of_laureates = n()) |>
  drop_na(Prize_year)

Nnnobel <- Nobel |>
  mutate(
    Cent = case_when(
      Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
      Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
      Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
      Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
      Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
    ),
    Cent = factor(Cent, levels = c("1901-1925","1926-1950",
                                   "1951-1975","1976-2000","2001-2025")),
    Category = factor(Category)
  ) |>
  drop_na(Category, Cent) |>
  count(Cent, Category)
```


Sidebar {.sidebar}
-------------------------------------------------------------

```{r, echo=FALSE}
selectInput("Cent1", label = "Time Period:",
            choices = unique(Nnnobel$Cent),
            selected = "1901-1925")
```

## Nobel Prize by Category (Interactive)

```{r, echo=FALSE}
renderPlot({
  Nnnobel |>
    filter(Cent == input$Cent1) |>
    ggplot(aes(x = Category, y = n)) +
    geom_col(fill = "black", colour = "#d6c089") +
    labs(title = paste("Nobel Prize Categories:", input$Cent1),
         x = "Category", y = "Number of Prizes") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```

Overview Continued
==========================


### Distribution of Nobel Laureates Over Time

```{r, echo=FALSE}
highchart() |>
  hc_add_series(
    data = Nnobel,
    type = "line",
    hcaes(x = Prize_year, y = number_of_laureates),
    name = "Number of Laureates"
  ) |>
  hc_title(text = "Nobel Laureates Over Time")|>
  hc_xAxis(categories = unique(Nnobel$Prize_year),
    title = list(text = "Year")) |>
  hc_yAxis(title = list(text="Number of Laureates")) |>
  hc_plotOptions(series = list(animation = TRUE)) |>
  hc_legend(align = "right", 
            verticalAlign = "top") |>
  hc_colors(c("#d6c089"))
```


### Laureate Age Distribution

```{r}
Nobel |>
  drop_na(Age) |>
  ggplot(aes(x = Age)) +
  geom_histogram(bins = 30, binwidth = 3.5, fill = "black", colour = "#d6c089") +
  labs(title = "Laureate Age Distribution",
       x = "Age", y = "Number of Laureates") +
  theme_minimal()
```


### Comments 
The scatterplot shows that the number of Nobel laureates generally increased over time, especially after the mid-1900s. This may reflect growing global participation and scientific development.

The histogram shows that most Nobel Prize laureates receive their awards between ages 50 and 75, especially around their early 60s. This suggests that Nobel recognition often comes after many years of achievement.


Geographic Analysis
==========================

Row {data-height=450}
-----------------------------------------------------------------------

### Chloropleth map

```{r}

# creating a country_count data set with names matching earth data
nobel_count <- Nobel |>
  filter(!is.na(Birth_country)) |>
  mutate(Birth_country = recode(Birth_country,
    "USA"                               = "United States of America",
    "the Netherlands"                   = "Netherlands",
    "South Korea"                       = "South Korea",
    "Korea"                             = "South Korea",
    "USSR"                              = "Russia",
    "Russian Empire"                    = "Russia",
    "Prussia"                           = "Germany",
    "Bavaria"                           = "Germany",
    "Mecklenburg"                       = "Germany",
    "Württemberg"                       = "Germany",
    "Hesse-Kassel"                      = "Germany",
    "West Germany"                      = "Germany",
    "East Friesland"                    = "Germany",
    "Schleswig"                         = "Germany",
    "Free City of Danzig"               = "Poland",
    "German-occupied Poland"            = "Poland",
    "Austria-Hungary"                   = "Austria",
    "Austrian Empire"                   = "Austria",
    "Ottoman Empire"                    = "Turkey",
    "Persia"                            = "Iran",
    "British Protectorate of Palestine" = "Israel",
    "British Mandate of Palestine"      = "Israel",
    "British India"                     = "India",
    "Burma"                             = "Myanmar",
    "Tuscany"                           = "Italy",
    "Gold Coast"                        = "Ghana",
    "Belgian Congo"                     = "Dem. Rep. Congo",
    "Southern Rhodesia"                 = "Zimbabwe",
    "French Algeria"                    = "Algeria",
    "French protectorate of Tunisia"    = "Tunisia",
    "Czechoslovakia"                    = "Czechia",
    "East Timor"                        = "Timor-Leste",
    "Java, Dutch East Indies"           = "Indonesia",
    "Faroe Islands (Denmark)"           = "Denmark",
    "Guadeloupe, France"                = "France",
    "British West Indies"               = "Trinidad and Tobago",
    "Crete"                             = "Greece",
    "Scotland"                          = "United Kingdom",
    "Northern Ireland"                  = "United Kingdom",
    "Tibet"                             = "China",
    "Bosnia"                            = "Bosnia and Herz."
  )) |>
  group_by(Birth_country) |>
  summarise(total = n()) # getting the counts for each country as total

# creating a data set with country coordinates and boundaries 
world <- ne_countries(scale = "medium", returnclass = "sf")

# Merging our country_count data set with the coordinate data set
world_nobel <- world |>
  left_join(nobel_count, by = c("name" = "Birth_country")) |>
  mutate(total = replace_na(total, 0)) #replacing nas with 0 

# selecting a color palette and assigning bins 
pal <- colorBin(
  palette  = "YlOrRd",
  domain   = world_nobel$total[world_nobel$total > 0],
  bins     = c(1, 5, 10, 25, 50, 100, 300),
  na.color = "#f0f0f0"
)

# Map 
leaflet(world_nobel) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = 0, lat = 20, zoom = 2) |>
  addPolygons(
    fillColor    = ~ifelse(total == 0, "#f0f0f0", pal(ifelse(total == 0, NA, total))),
    fillOpacity  = ~ifelse(total == 0, 0.3, 0.8),
    color        = "white",
    weight       = 0.5,
    label        = ~paste0(name, ": ", ifelse(total == 0, "None", paste0(total, " Nobel Prize(s)"))),# saying none for countries without prizes
    highlight    = highlightOptions(
      weight       = 3,
      color        = "#333",
      bringToFront = TRUE
    )
  ) |>
  addLegend(
    pal      = pal,
    values   = world_nobel$total[world_nobel$total > 0],
    title    = "Nobel Prizes",
    position = "bottomright",
    labFormat = labelFormat(suffix = " prizes")
  )
```


Geographical Analysis Continued
==========================

```{r}

Nnnobel_1 <- Nobel |>
  filter(!is.na(Birth_country)) |>
  mutate(Birth_country = recode(Birth_country,
    "USA"                               = "United States of America",
    "the Netherlands"                   = "Netherlands",
    "South Korea"                       = "South Korea",
    "Korea"                             = "South Korea",
    "USSR"                              = "Russia",
    "Russian Empire"                    = "Russia",
    "Prussia"                           = "Germany",
    "Bavaria"                           = "Germany",
    "Mecklenburg"                       = "Germany",
    "Württemberg"                       = "Germany",
    "Hesse-Kassel"                      = "Germany",
    "West Germany"                      = "Germany",
    "East Friesland"                    = "Germany",
    "Schleswig"                         = "Germany",
    "Free City of Danzig"               = "Poland",
    "German-occupied Poland"            = "Poland",
    "Austria-Hungary"                   = "Austria",
    "Austrian Empire"                   = "Austria",
    "Ottoman Empire"                    = "Turkey",
    "Persia"                            = "Iran",
    "British Protectorate of Palestine" = "Israel",
    "British Mandate of Palestine"      = "Israel",
    "British India"                     = "India",
    "Burma"                             = "Myanmar",
    "Tuscany"                           = "Italy",
    "Gold Coast"                        = "Ghana",
    "Belgian Congo"                     = "Dem. Rep. Congo",
    "Southern Rhodesia"                 = "Zimbabwe",
    "French Algeria"                    = "Algeria",
    "French protectorate of Tunisia"    = "Tunisia",
    "Czechoslovakia"                    = "Czechia",
    "East Timor"                        = "Timor-Leste",
    "Java, Dutch East Indies"           = "Indonesia",
    "Faroe Islands (Denmark)"           = "Denmark",
    "Guadeloupe, France"                = "France",
    "British West Indies"               = "Trinidad and Tobago",
    "Crete"                             = "Greece",
    "Scotland"                          = "United Kingdom",
    "Northern Ireland"                  = "United Kingdom",
    "Tibet"                             = "China",
    "Bosnia"                            = "Bosnia and Herz."
  )) |>
  # join continent from world_nobel (drop geometry to keep it a plain data frame)
  left_join(
    world_nobel |> st_drop_geometry() |> select(name, continent),
    by = c("Birth_country" = "name")
  ) |>
  mutate(Cent = case_when(
    Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
    Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
    Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
    Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
    Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
  )) |>
  drop_na(continent, Cent) |>
  count(Cent, continent, name = "n")
```


Sidebar {.sidebar}
-------------------------------------------------------------


```{r, echo=FALSE}
selectInput("Cent", label = "Time Period:",
            choices = unique(Nnnobel_1$Cent),
            selected = "1901-1925")
```



## Continent

```{r, echo=FALSE}
renderPlot({
  Nnnobel_1 |>
    filter(Cent == input$Cent) |>
    ggplot(aes(x = continent, y = n)) +
    geom_col(fill = "black", colour = "#d6c089") +
    labs(title = paste("Nobel Prizes by Continent:", input$Cent),
         x = "Continent", y = "Number of Prizes") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```

## Comments 

  Across most historical periods, Western countries dominated Nobel Prize awards, showing how scientific prestige and opportunities have not been equally distributed worldwide.

Gender analysis
==========================

### Gender Overall

```{r, echo=FALSE}
Nobel_6 <- Nobel |>
  filter(!is.na(Gender)) |>
  count(Gender) -> Gender

highchart() |> 
  hc_chart(type ="column",
           options3d = list(enabled = TRUE, beta = 15, alpha = 15)) |>
  hc_xAxis(categories = unique(Nobel_6$Gender),
    title = list(text = "Gender")) |>
  hc_add_series(data = Gender$n, name = "Gender") |>
  hc_colors(c("#d6c089"))
```

### Gender by Era

```{r, echo=FALSE}
Nobel_3 <- Nobel |> 
  mutate(Cent = case_when(
    Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
    Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
    Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
    Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
    Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
  )) |>
  drop_na(Gender) |>
  count(Cent, Gender)

highchart() |>
  hc_add_series(data = Nobel_3,
                type = "column", 
                hcaes(x = Cent, y = n, group = Gender)) |>
  hc_xAxis(categories = unique(Nobel_3$Cent),
    title = list(text = "Year")) |>
  hc_yAxis(title = list(text="Number of Laureates")) |>
  hc_plotOptions(series = list(animation = TRUE)) |>
  hc_legend(align = "right", verticalAlign = "top") |>
  hc_colors(c("#d6c089", "black"))
```

### Gender by Category

```{r}
Nobel_10 <- Nobel |>
  drop_na(Category, Gender) |>
  count(Category, Gender)

highchart() |>
  hc_add_series(
    data = Nobel_10,
    type = "column",
    hcaes(x = Category, y = n, group = Gender)
  ) |>
  hc_plotOptions(
    column = list(stacking = "percent")
  ) |>
  hc_yAxis(
    title = list(text = "Percentage of Laureates")
  ) |>
  hc_xAxis(categories = unique(Nobel_10$Category),
    title = list(text = "Category")
  ) |>
  hc_colors(c("#d6c089", "black"))
```


### Comments {data-width=300}
Nobel Prize awards have historically been dominated by male laureates, while female winners represent a much smaller share of total recipients.

Female representation gradually increased across time periods, especially after 1975, showing slow progress toward greater gender inclusion.

Peace and Literature show higher female representation, while categories such as Physics and Economics remain heavily male-dominated.


Conclusion
==========================

In conclusion , the number of Nobel laureates has increased over time, reflecting a shift toward group work and recognition. While early Nobel Prizes were concentrated in categories like Physics, , the distribution has gradually become more balanced, indicating a broader appreciation of contributions across disciplines. The age distribution further shows that Nobel Prizes are typically awarded later in life, highlighting the long-term nature of impactful work and the delay between discovery and recognition.


However, significant disparities remain. The gender analysis demonstrates a strong historical imbalance, with male laureates overwhelmingly dominating total counts. Although female representation has increased in more recent decades, progress has been slow and uneven across categories, with scientific fields still showing the greatest gender gaps. This suggests that gender inequality in Nobel recognition is shaped by both historical exclusion and structural barriers within certain disciplines.


Geographically, Nobel Prizes are highly concentrated in a small number of countries, particularly in North America and Western Europe. Large regions of the world remain underrepresented, reflecting global inequalities in access to education, research funding, and institutional support. Additionally, because the data is based on birth country, these patterns also capture migration trends and the movement of talent toward well-resourced regions.


Altogether, these findings show that while the Nobel Prize has evolved to reflect changes in science and society, it continues to be influenced by broader social, historical, and structural factors which is reflected in the disparities.

 