---
title: "The Nobel Prize"
output: 
  flexdashboard::flex_dashboard:
    css: styles.css
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r global, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(ggplot2)
library(highcharter)

Nobel <- read.csv("Nobel_final_data.csv")
```

Introduction
==========================

#### Background Information

The Nobel Prize is one of the most prestigious international awards, recognizing individuals and organizations that have made major contributions to science, literature, economics, and peace. 

In this Shiny dashboard, we analyze Nobel Prize data collected from the official Nobel Prize API. The dataset includes information about laureates such as age, gender, birth country, prize category, prize amount, and year awarded.

Using this project, we explore historical trends in Nobel Prize awards across different time periods and categories. We also investigate patterns related to gender representation and geographical distribution in order to better understand inequalities and changes in recognition over time.



#### Data Sources

- [Nobel Prize Website](https://www.nobelprize.org/)
- [Nobel Prize API Page](https://www.nobelprize.org/about/developer-zone-2/)
- [Nobel Prize API](https://api.nobelprize.org/2.1/laureates)

#### GitHub

Here's a link to our [code](https://github.com/)
 
Row {data-width=6}
-------------------------------------

Column {data-width=6}
-------------------------------------

- Rosalind Franklin

- Lise Meitner

- Alice Ball 

- Jocelyn Bell Burnell

- Marthe Gautier

- Cecilia Payne-Gaposchkin

- Esther Lederberg 

- Chien-Shiung Wu

- Margaret Knight

- Hedy Lamarr

- Ada Lovelace

 
Overview
==========================

```{r, include=FALSE}
library(shiny)
library(dplyr)
library(ggplot2)
library(flexdashboard)
library(highcharter)

Nnobel <- Nobel |>
  group_by(Prize_year) |>
  summarise(number_of_laureates = n()) |>
  drop_na(Prize_year)

Nnnobel <- Nobel |>
  mutate(
    Cent = case_when(
      Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
      Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
      Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
      Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
      Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
    ),
    Cent = factor(Cent, levels = c("1901-1925","1926-1950",
                                   "1951-1975","1976-2000","2001-2025")),
    Category = factor(Category)
  ) |>
  drop_na(Category, Cent) |>
  count(Cent, Category)
```


Sidebar {.sidebar}
-------------------------------------------------------------

```{r}
selectInput("Cent", label = "Time Period:",
            choices = unique(Nnnobel$Cent),
            selected = "1901-1925")
```

## Nobel Prize Categories (Interactive)

```{r}
renderPlot({
  Nnnobel |>
    filter(Cent == input$Cent) |>
    ggplot(aes(x = Category, y = n)) +
    geom_col(fill = "black", colour = "gold") +
    labs(title = paste("Nobel Prize Categories:", input$Cent),
         x = "Category", y = "Number of Prizes") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```


## Distribution of Nobel Laureates Over Time

```{r, echo=FALSE}
ggplot(Nnobel, aes(x = Prize_year, y = number_of_laureates)) +
  geom_point(colour = "black") +
  geom_smooth(se = FALSE, colour = "gold") +
  labs(title = "Distribution of Nobel Laureates Over Time",
       x = "Year", y = "Count") +
  theme_minimal()
```


### Laureate Age Distribution

```{r}
Nobel |>
  drop_na(Age) |>
  ggplot(aes(x = Age)) +
  geom_histogram(bins = 30, binwidth = 3.5, fill = "black", colour = "gold") +
  labs(title = "Laureate Age Distribution",
       x = "Age", y = "Number of Laureates") +
  theme_minimal()
```


## Comments {data-width=300}

let us see if this works or not. god safe us 


Geographic Analysis
==========================

Row {data-height=450}
-----------------------------------------------------------------------

### yeah

```{r}
library(leaflet)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(sf)
library(tidyr)

# creating a country_count data set with names matching earth data
nobel_count <- Nobel |>
  filter(!is.na(Birth_country)) |>
  mutate(Birth_country = recode(Birth_country,
    "USA"                               = "United States of America",
    "the Netherlands"                   = "Netherlands",
    "South Korea"                       = "South Korea",
    "Korea"                             = "South Korea",
    "USSR"                              = "Russia",
    "Russian Empire"                    = "Russia",
    "Prussia"                           = "Germany",
    "Bavaria"                           = "Germany",
    "Mecklenburg"                       = "Germany",
    "Württemberg"                       = "Germany",
    "Hesse-Kassel"                      = "Germany",
    "West Germany"                      = "Germany",
    "East Friesland"                    = "Germany",
    "Schleswig"                         = "Germany",
    "Free City of Danzig"               = "Poland",
    "German-occupied Poland"            = "Poland",
    "Austria-Hungary"                   = "Austria",
    "Austrian Empire"                   = "Austria",
    "Ottoman Empire"                    = "Turkey",
    "Persia"                            = "Iran",
    "British Protectorate of Palestine" = "Israel",
    "British Mandate of Palestine"      = "Israel",
    "British India"                     = "India",
    "Burma"                             = "Myanmar",
    "Tuscany"                           = "Italy",
    "Gold Coast"                        = "Ghana",
    "Belgian Congo"                     = "Dem. Rep. Congo",
    "Southern Rhodesia"                 = "Zimbabwe",
    "French Algeria"                    = "Algeria",
    "French protectorate of Tunisia"    = "Tunisia",
    "Czechoslovakia"                    = "Czechia",
    "East Timor"                        = "Timor-Leste",
    "Java, Dutch East Indies"           = "Indonesia",
    "Faroe Islands (Denmark)"           = "Denmark",
    "Guadeloupe, France"                = "France",
    "British West Indies"               = "Trinidad and Tobago",
    "Crete"                             = "Greece",
    "Scotland"                          = "United Kingdom",
    "Northern Ireland"                  = "United Kingdom",
    "Tibet"                             = "China",
    "Bosnia"                            = "Bosnia and Herz."
  )) |>
  group_by(Birth_country) |>
  summarise(total = n()) # getting the counts for each country as total

# creating a data set with country coordinates and boundaries 
world <- ne_countries(scale = "medium", returnclass = "sf")

# Merging our country_count data set with the coordinate data set
world_nobel <- world |>
  left_join(nobel_count, by = c("name" = "Birth_country")) |>
  mutate(total = replace_na(total, 0)) #replacing nas with 0 

# selecting a color palette and assigning bins 
pal <- colorBin(
  palette  = "YlOrRd",
  domain   = world_nobel$total[world_nobel$total > 0],
  bins     = c(1, 5, 10, 25, 50, 100, 300),
  na.color = "#f0f0f0"
)

# Map 
leaflet(world_nobel) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = 0, lat = 20, zoom = 2) |>
  addPolygons(
    fillColor    = ~ifelse(total == 0, "#f0f0f0", pal(ifelse(total == 0, NA, total))),
    fillOpacity  = ~ifelse(total == 0, 0.3, 0.8),
    color        = "white",
    weight       = 0.5,
    label        = ~paste0(name, ": ", ifelse(total == 0, "None", paste0(total, " Nobel Prize(s)"))),# saying none for countries without prizes
    highlight    = highlightOptions(
      weight       = 3,
      color        = "#333",
      bringToFront = TRUE
    )
  ) |>
  addLegend(
    pal      = pal,
    values   = world_nobel$total[world_nobel$total > 0],
    title    = "Nobel Prizes",
    position = "bottomright",
    labFormat = labelFormat(suffix = " prizes")
  )
```


more geography
==========================

Sidebar {.sidebar}
-------------------------------------------------------------

```{r}
selectInput("Cent", label = "Time Period:",
            choices = unique(Nnnobel$Cent),
            selected = "1901-1925")
```


### continent

```{r}
Nnnobel <- Nobel |>
  filter(!is.na(Birth_country)) |>
  mutate(Birth_country = recode(Birth_country,
    "USA"                               = "United States of America",
    "the Netherlands"                   = "Netherlands",
    "South Korea"                       = "South Korea",
    "Korea"                             = "South Korea",
    "USSR"                              = "Russia",
    "Russian Empire"                    = "Russia",
    "Prussia"                           = "Germany",
    "Bavaria"                           = "Germany",
    "Mecklenburg"                       = "Germany",
    "Württemberg"                       = "Germany",
    "Hesse-Kassel"                      = "Germany",
    "West Germany"                      = "Germany",
    "East Friesland"                    = "Germany",
    "Schleswig"                         = "Germany",
    "Free City of Danzig"               = "Poland",
    "German-occupied Poland"            = "Poland",
    "Austria-Hungary"                   = "Austria",
    "Austrian Empire"                   = "Austria",
    "Ottoman Empire"                    = "Turkey",
    "Persia"                            = "Iran",
    "British Protectorate of Palestine" = "Israel",
    "British Mandate of Palestine"      = "Israel",
    "British India"                     = "India",
    "Burma"                             = "Myanmar",
    "Tuscany"                           = "Italy",
    "Gold Coast"                        = "Ghana",
    "Belgian Congo"                     = "Dem. Rep. Congo",
    "Southern Rhodesia"                 = "Zimbabwe",
    "French Algeria"                    = "Algeria",
    "French protectorate of Tunisia"    = "Tunisia",
    "Czechoslovakia"                    = "Czechia",
    "East Timor"                        = "Timor-Leste",
    "Java, Dutch East Indies"           = "Indonesia",
    "Faroe Islands (Denmark)"           = "Denmark",
    "Guadeloupe, France"                = "France",
    "British West Indies"               = "Trinidad and Tobago",
    "Crete"                             = "Greece",
    "Scotland"                          = "United Kingdom",
    "Northern Ireland"                  = "United Kingdom",
    "Tibet"                             = "China",
    "Bosnia"                            = "Bosnia and Herz."
  )) |>
  # join continent from world_nobel (drop geometry to keep it a plain data frame)
  left_join(
    world_nobel |> st_drop_geometry() |> select(name, continent),
    by = c("Birth_country" = "name")
  ) |>
  mutate(Cent = case_when(
    Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
    Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
    Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
    Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
    Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
  )) |>
  drop_na(continent, Cent) |>
  count(Cent, continent, name = "n")
```

```{r}
renderPlot({
  Nnnobel |>
    filter(Cent == input$Cent) |>
    ggplot(aes(x = continent, y = n)) +
    geom_col(fill = "black", colour = "gold") +
    labs(title = paste("Nobel Prizes by Continent:", input$Cent),
         x = "Continent", y = "Number of Prizes") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```


Comments {data-width=300}


let us see if this works or not. god safe us 


Gender analysis
==========================

### Gender Overall

```{r, echo=FALSE}
Nobel_6 <- Nobel |>
  filter(!is.na(Gender)) |>
  count(Gender) -> Gender

highchart() |> 
  hc_chart(type ="column",
           options3d = list(enabled = TRUE, beta = 15, alpha = 15)) |>
  hc_xAxis(categories = unique(Nobel_6$Gender),
    title = list(text = "Gender")) |>
  hc_add_series(data = Gender$n, name = "Gender") |>
  hc_colors(c("#d6c089"))
```

### Gender by Era

```{r, echo=FALSE}
library(dplyr)
Nobel_3 <- Nobel |> 
  mutate(Cent = case_when(
    Prize_year >= 1901 & Prize_year <= 1925 ~ "1901-1925",
    Prize_year >= 1926 & Prize_year <= 1950 ~ "1926-1950",
    Prize_year >= 1951 & Prize_year <= 1975 ~ "1951-1975",
    Prize_year >= 1976 & Prize_year <= 2000 ~ "1976-2000",
    Prize_year >= 2001 & Prize_year <= 2025 ~ "2001-2025"
  )) |>
  drop_na(Gender) |>
  count(Cent, Gender)

highchart() |>
  hc_add_series(data = Nobel_3,
                type = "column", 
                hcaes(x = Cent, y = n, group = Gender)) |>
  hc_xAxis(categories = unique(Nobel_3$Cent),
    title = list(text = "Year")) |>
  hc_yAxis(title = list(text="Number of Laureates")) |>
  hc_plotOptions(series = list(animation = TRUE)) |>
  hc_legend(align = "right", verticalAlign = "top") |>
  hc_colors(c("#d6c089"))
```

### Gender by Category

```{r}
Nobel_10 <- Nobel |>
  drop_na(Category, Gender) |>
  count(Category, Gender)

highchart() |>
  hc_add_series(
    data = Nobel_10,
    type = "column",
    hcaes(x = Category, y = n, group = Gender)
  ) |>
  hc_plotOptions(
    column = list(stacking = "percent")
  ) |>
  hc_yAxis(
    title = list(text = "Percentage of Laureates")
  ) |>
  hc_xAxis(categories = unique(Nobel_10$Category),
    title = list(text = "Category")
  )
```


### Comments {data-width=300}


let us see if this works or not. god safe us 